Type-Safe Modular Parsing

### 2016-12-20 paper/sections/Overview.tex
Rewrite, add references

Although there are many parsing techniques, not all of them are
suitable for type-safe modular parsing. In particular there are many
techniques which fail to provide modular type-checking and separate
compilation. Moreover, even if modular type-checking and separate
compilation are supported, efficiency is another
concern. A parsing technique should have low overhead when applied
in a modular setting. In the remaider of this section, we eliminate
various techniques that fail to satisfy our requirements, and argue
that that Packrat parsing~\cite{} is a suitable candidate for
type-safe modular parsing.

\paragraph{Parser Generators} The most widely use tools for parsing
are parser generators~\cite{}. Parser generators help users generate parsers automatically or
semi-automatically from a given grammar. There is no restriction on
the algorithm, while most of them adopts table-based LL~\cite{} and LR parsing
algorithms~\cite{}.
Although efficient, the main drawback of parser generators is that they do not support
modular type-checking and separate compilation.

Modular parsing based on parser generators is supported by many libraries []. Users can separate the concrete/abstract syntax and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA [] uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence those libraries only have syntactic modularity.

@@

\huang{a paragraph here, refer to related work section, discuss
  previous works of modular parsing using parser generators and its
  drawbacks}\bruno{Please finish this.}\huang{done}

\paragraph{Parser Combinators}
Comparing with the parser generators, a \textit{parser combinator}~\cite{}
takes several parsers and produce a new parser as its output. Parser combinators are
popular in functional programming, where the parsers are represented
by functions and parser combinators are higher-order functions accepts
them.

@@

Unfortunately many parser combinators have important limitations for
modular parsing. In particular several parser combinators,
including the famous library Parsec~\cite{} library, require
programmers to manually do \textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

@@

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. Take Parsec as an example, its choice
combinator only tries the second alternative if the first fails
without any token consumption. If two alternatives have a same
non-empty prefix, an auxiliary function \inlinecode{try} must be added
to backtrack. However, if we want to extend our parsers later, we
should always consider the worst case in which all alternatives share
common prefixes. Then we need to add \inlinecode{try} for all the
branches, which results worst case exponential time complexity.
\huang{done}\bruno{you should again illustrate your point here with an
  example. The example should illustrate that the addition of a new
  case may require introducing backtracking. }

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

Unfortunately, this extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \inlinecode{(\x.x)} \inlinecode{(\y.y)} which is obviously valid.

@@

That is the reason why the extended parser fails to parse some valid expressions such as \inlinecode{(\x.x)} \inlinecode{(\y.y)}. To resolve this issue, we need to let \inlinecode{pApp} be aware of the change of grammar. We have some direct solutions:

@@

\paragraph{Client Code} Since the type of parsing functions are
changed, we can not apply them directly. The client code below shows
how to use the new parser. The \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. \bruno{Too much detail is being given here about open
  recursion. In Section 2 we should talk only about \emph{how} to use 
open recursion, and not how open recursion is implemented. We 
can summarize, for example the API related to open recursion. 
Section 4 is where the implementation and detailed explanation 
about open recursion should come.
}

\lstinputlisting[linerange=68-75]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_USE

@@

\lstinputlisting[linerange=80-86]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

@@

\lstinputlisting[linerange=95-95]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

@@

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

@@

To use the parser, we must provide an algebra instance as the operation to construct the parsing result. As shown in the code below, we feed the pretty printing operation \inlinecode{ExprPrint} to the generic \inlinecode{parse} function. It will print \inlinecode{(x y)} as the result.

\lstinputlisting[linerange=38-43]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

@@

\lstinputlisting[linerange=48-62]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT

@@

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can use this new operation during parsing.

\lstinputlisting[linerange=75-81]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

@@

\lstinputlisting[linerange=66-70]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX

>>>>

Although there are many parsing techniques, not all of them are
suitable for type-safe modular parsing. In particular there are many
techniques which fail to provide modular type-checking and separate
compilation. Moreover, even if modular type-checking and separate
compilation are supported, efficiency is another
concern. A parsing technique should have low overhead when applied
in a modular setting. In the remaider of this section, we eliminate
various techniques that fail to satisfy our requirements, and argue
that that Packrat parsing~\cite{Ford2002} is a suitable candidate for
type-safe modular parsing.

\paragraph{Parser Generators} The most widely use tools for parsing
are parser generators. Parser generators help users generate parsers automatically or
semi-automatically from a given grammar. There is no restriction on
the algorithm, while most of them adopts table-based LL~\cite{lewis1968syntax} and LR~\cite{knuth1965translation} parsing
algorithms.
Although efficient, the main drawback of parser generators is that they do not support
modular type-checking and separate compilation.

Modular parsing based on parser generators is supported by many libraries~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}. Users can separate the syntax definition and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA~\cite{Gouseti2014} uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence those libraries only have syntactic modularity.

@@

\paragraph{Parser Combinators}
Comparing with the parser generators, a \textit{parser combinator}~\cite{burge1975,Wadler1985}
takes several parsers and produce a new parser as its output. Parser combinators are
popular in functional programming, where the parsers are represented
by functions and parser combinators are higher-order functions accepts
them.

@@

Unfortunately many parser combinators have important limitations for
modular parsing. In particular several parser combinators,
including the famous library Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

@@

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. Take Parsec as an example, its choice
combinator only tries the second alternative if the first fails
without any token consumption. If two alternatives have a same
non-empty prefix, an auxiliary function \inlinecode{try} must be added
to backtrack. However, if we want to extend our parsers later, we
should always consider the worst case in which all alternatives share
common prefixes. Then we need to add \inlinecode{try} for all the
branches, which results worst case exponential time complexity.

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly~\cite{warth2008}. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

Unfortunately, this extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \inlinecode{(\\x.x)} \inlinecode{(\\y.y)} which is obviously valid.

@@

That is the reason why the extended parser fails to parse some valid expressions such as \inlinecode{(\\x.x)} \inlinecode{(\\y.y)}. To resolve this issue, we need to let \inlinecode{pApp} be aware of the change of grammar. We have some direct solutions:

@@

\paragraph{Client Code} Since the type of parsing functions are
changed, we can not apply them directly. A \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. After that, it can be used as an ordinary parser.
The client code below demonstrates the usage of new parsers.

\huang{done}\bruno{Too much detail is being given here about open
  recursion. In Section 2 we should talk only about \emph{how} to use
open recursion, and not how open recursion is implemented. We
can summarize, for example the API related to open recursion.
Section 4 is where the implementation and detailed explanation
about open recursion should come.
}

\lstinputlisting[linerange=68-72]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_USE

@@

\lstinputlisting[linerange=77-83]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

@@

\lstinputlisting[linerange=89-89]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

@@

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{Oliveira2012} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

@@

@@

\lstinputlisting[linerange=32-46]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT

@@

To use the parser, we must provide an algebra instance as the operation to construct the parsing results. In the code below, we use the pretty printing operation \inlinecode{LambdaPrint} as the algebra instance, so that the parsing result is a pretty printed AST.

\lstinputlisting[linerange=55-60]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can use this operation during parsing and obtain a set of free variables.

\lstinputlisting[linerange=74-80]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

@@

\lstinputlisting[linerange=65-69]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
